---
title: "Storm analysis"
author: "Koen Vermeulen"
date: "7-12-2019"
output: 
        html_document:
                keep_md: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = T, 
                      cache = T)
```

# Summary
This report contains a storm analysis. It answers two questions: 
1. Across the United States, which types of events are most harmful with respect to population health?
2. Across the United States, which types of events have the greatest economic consequences?
After some data processing, the questions are answered by two bar chart visualisations. The most harmful type of events shows to be tornado's. The event type with the most economic damages is a flood.

## Loading necessary libraries

```{r libraries, warning=FALSE}
library(dplyr)
library(stringr)
library(lubridate)
library(ggplot2)
```

## Data Processing

### Downloading the data

If the data is not already in the working directory, it is downloaded from the correct URL.

```{r downloading, warning=FALSE}
filename <- 'storms.csv.bz2'
if (!file.exists(filename)) {
  download.file('https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2', filename)
}
storms <- read.csv(filename)
```

### Correcting dates

In the documentation on the dataset it was mentioned that the data was properly collected after 1996. This report focussed on data from 2000 and later. Therefore it was necessary to properly store the BGN_DATE variable. 

```{r dates}
storms <- storms %>%
        mutate(begin_date_char = as.character(storms$BGN_DATE)) %>%
        mutate(begin_date_nchar = nchar(begin_date_char)) %>%
        mutate(begin_date = as.Date(str_sub(begin_date_char, 
                                            start = 1,
                                            end = begin_date_nchar - 8), 
                                    tryFormats = "%m/%d/%Y")) %>%
        mutate(begin_date_year = year(begin_date))
```

### Reducing data to recent times

After correcting the date variable it was possible to filter the data to 2000 and later.

```{r}
# Reducing data to recent times (from 2000)
storms2000 <- filter(storms, begin_date_year >= 2000)
range(storms2000$begin_date)
sapply(storms2000, function(x) sum(is.na(x)))
```

### Revalue and combining damages & combining harm

The damages (property and crop) were reported in the dataset in seperate columns: 
1. PROPDMG entered as actual dollar amount.
2. PROPDMGEXP containing the magnitude ("K" for thousands, "M" for millions & "B" for billions)
In order to analyse the damages the damages were multiplied by the magnitude for both property and crop damages. After that the two total columns were summed to assess the total economic impact.

The harmful impact on the population was reported in the dataset by two variables, INJURIES & FATALITIES. In order to analyse the harm inflicted by the types of events, the two variables were added together. 

```{r}
storms2000dmg <- storms2000 %>%
        mutate(PROPDMGEXPnum = 1) %>%
        mutate(CROPDMGEXPnum = 1) %>%
        mutate(PROPDMGEXPnum = ifelse(PROPDMGEXP == "K", 1000, PROPDMGEXPnum)) %>%
        mutate(CROPDMGEXPnum = ifelse(CROPDMGEXP == "K", 1000, CROPDMGEXPnum)) %>%
        mutate(PROPDMGEXPnum = ifelse(PROPDMGEXP == "M", 1000000, PROPDMGEXPnum)) %>%
        mutate(CROPDMGEXPnum = ifelse(CROPDMGEXP == "M", 1000000, CROPDMGEXPnum)) %>%
        mutate(PROPDMGEXPnum = ifelse(PROPDMGEXP == "B", 1000000000, PROPDMGEXPnum)) %>%
        mutate(CROPDMGEXPnum = ifelse(CROPDMGEXP == "B", 1000000000, CROPDMGEXPnum)) %>%
        mutate(PROPDMGtotal = PROPDMG * PROPDMGEXPnum) %>%
        mutate(CROPDMGtotal = CROPDMG * CROPDMGEXPnum) %>%
        mutate(DMGtotal = PROPDMGtotal + CROPDMGtotal) %>%
        mutate(HARMEDtotal = INJURIES + FATALITIES)

```

### Grouping damages and harming impact by event type

To assess damages and harm in relation to the types of events, the dataset was grouped by the event type. For the damages and harm the sum was summarized in this group by.

```{r grouping by type}
types <- storms2000dmg %>%
        group_by(EVTYPE) %>%
        summarize(DMG_SUM = sum(DMGtotal),
                  HARMED_SUM = sum(HARMEDtotal))
```

## Results

### Question 1: Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?

The barchart below shows that tornados are the most harmful with respect to the population health.

```{r results question 1}
harm <- types %>%
        top_n(n = 5, wt = HARMED_SUM) %>%
        ggplot(aes(y = HARMED_SUM, x = reorder(x = EVTYPE, X = HARMED_SUM), fill=EVTYPE)) +
        geom_bar(stat = "identity", show.legend = FALSE) +
        labs(title = "Top 5 most harmful event types", 
             x = "", y = "Number of harmed people involved") +
        coord_flip() + 
        theme_light()
harm
```

### Question 2: Across the United States, which types of events have the greatest economic consequences?

The barchart below shows that floods have the greatest economic consequences. 

```{r results question 2}
damage <- types %>%
        top_n(n = 5, wt = DMG_SUM) %>%
        ggplot(aes(y = DMG_SUM, x = reorder(x = EVTYPE, X = DMG_SUM), fill=EVTYPE)) +
        geom_bar(stat = "identity", show.legend = FALSE) +
        labs(title = "Top 5 event types with the greatest economic consequences",
             x = "", y = "Total economic damages") +
        coord_flip() +
        theme_light()
damage
```
